\section{Related and Further Work}

Our best effort has gone into ensuring the correctness of our pen-and-paper
soundness proof for DIF, and given similar soundness results and the success of
Scala, it seems likely that our soundness theorem holds in truth. However, a
mechanised proof would give a higher level of confidence in this result.

It is known that type assignment and subtyping are undecidable in DOT
\cite{AR16} since DOT can encode system $F_{<:}$. DOT has no principal types. As
DIF is a superset of DOT, these properties hold for DIF transitively. DOT does,
however, have a local type inference procedure \cite{PT00}. It follows that such
a procedure should also exist for DIF, especially since local type inference is
used to great effect in Scala, a language with implicits.

It is widely thought that there is a correspondence between implicits and type
classes - it is easily seen that the dictionary passing of type classes is
simulated with an implicit parameter. It would be interesting to establish this
correspondence formally. One can envision a translation $t_{c \rightarrow i}$
from type classes to implicits, and given the established results of type
classes to lambda calculus $t_{c \rightarrow l}$ \cite{WB89} and implicits to
lambda calculus $t_{i \rightarrow l}$ \cite{OBLB18}, it seems that it should be
possible to establish $t_{c \rightarrow i} \cdot t_{i \rightarrow l} \equiv t_{c
\rightarrow l}$. for some notion of $\equiv$.

Implicit program constructs for session-typed concurrency have been studied in
\cite{JB18}. While Scala concurrency libraries leverage implicits, it is unclear
whether their usage is orthogonal to concurrency, or whether there are deeper
connections between the two concepts that can be further studied in theory. This
is a topic for further study.
